---
title: "HCHS SES Index preprocessing n=17.000"
output:
  html_document: default
date: "2024-11-16"
authors: A. Klimesch, L. Ascone
references: ChatGPT was used in the development or adaptation of this code. OpenAI ChatGPT (November 2024 version), https://chat.openai.com/
dataset: initial 17.000 cases of HCHS "HCHS_Data_SPSS_labelled_sozio"
purpose: "this code prepares the variables in the HCHS dataset for calculating the SES-Index"
---

packages
```{r, results = FALSE, message=FALSE}
library(foreign) # for read.spss
library(dplyr) 
library(stringr) # for str_trim
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load dataset and duplicate
```{r}
hchs_ses_17000 <- read.csv("C:/Users/annek/Documents/PhD/PTSD/HCHS/SES_Index_paper/17000_sample_ses_index/HCHS_SES_raw_data_17000.csv")
```

Select required variables
```{r}
hchs_ses <- hchs_ses_17000[, c("DisclosureID", "HCH_SVAGE0001" , "HCH_SVSEX0001", # these two are needed to inform the imputation later
  "am_schube_ausb_001",
  "am_schube_ausb_003.A1",
  "am_schube_ausb_003.A2",
  "am_schube_ausb_003.A3",
  "am_schube_ausb_003.A4",
  "am_schube_ausb_003.A5",
  "am_schube_ausb_003.A6",
  "am_schube_ausb_003.A7",
  "am_schube_ausb_003.A8",
  "am_schube_ausb_003.A9",
  "am_schube_ausb_003.A10",
  "am_schube_ausb_003.A11",
  "am_schube_ausb_003.A12",
  "am_schube_ausb_023",
  "am_schube_ausb_024",
  "am_schube_ausb_025",
  "am_schube_ausb_026",
  "am_schube_ausb_027",
  "am_schube_ausb_028",
  "am_schube_ausb_029",
  "am_schube_ausb_030",
  "am_schube_ausb_037",
  "am_schube_ausb_038",
  "sozio_fami_006",
  "sozio_fami_007")]
```


Trim spaces in the values of all variables used in this script to avoid bugs caused by random spaces
```{r}
columns_to_trim <- c( # trim spaces in variable names to avoid introduction of NAs by coercion 
  "am_schube_ausb_001",
  "am_schube_ausb_003.A1",
  "am_schube_ausb_003.A2",
  "am_schube_ausb_003.A3",
  "am_schube_ausb_003.A4",
  "am_schube_ausb_003.A5",
  "am_schube_ausb_003.A6",
  "am_schube_ausb_003.A7",
  "am_schube_ausb_003.A8",
  "am_schube_ausb_003.A9",
  "am_schube_ausb_003.A10",
  "am_schube_ausb_003.A11",
  "am_schube_ausb_003.A12",
  "am_schube_ausb_023",
  "am_schube_ausb_024",
  "am_schube_ausb_025",
  "am_schube_ausb_026",
  "am_schube_ausb_027",
  "am_schube_ausb_028",
  "am_schube_ausb_029",
  "am_schube_ausb_030",
  "am_schube_ausb_037",
  "am_schube_ausb_038",
  "sozio_fami_006",
  "sozio_fami_007")
hchs_ses <- hchs_ses %>%
  mutate(across(all_of(columns_to_trim), str_trim))
```

###Education
##School Education
variable: am_schube_ausb_001
```{r}
table(hchs_ses$am_schube_ausb_001, useNA = "always") # frequency table of original variable 
```
The variable has 4 missing or unclassified values: NA, "10" (other/ non-specified), "8888" (don't want to respond), "9999" (don't know). Category 10 has to be classified based on the information in variable am_schube_ausb_002, the others are classified as NAs. 
Before going on with the code, manually assign an adequate value to the cases of the variable am_schube_ausb_001 with current value "10" (other/ non-specified) using the information provided in variable am_schube_ausb_002. Otherwise, classify them as NAs. 

Unify the missing values of variable am_schube_ausb_001 (in this case, we classify those with '10' as NAs as well)
```{r}
hchs_ses <- hchs_ses %>%
  mutate(school_edu = ifelse(am_schube_ausb_001 %in% c('8888', '9999', 'NA', '10'), NA_character_, am_schube_ausb_001))

table(hchs_ses$school_edu, useNA = "always") # frequency table of new variable for comparison with original frequency output 
```
The frequency table shows that the calssification of NAs was successful. All frequencies remain the same, except the NA variable. 

##Professional Education
Before going on with the code, manually assign an adequate value to the cases of the variable am_schube_ausb_003 with current value "12" (other educational degree) using the information provided by the participant. Otherwise, classify them as NAs (which we do here). 
For education, there are 11 variables, which need to be aggregated into one and the highest value per case is chosen as value for the new variable am_schube_ausb_003  
```{r}
table(hchs_ses$am_schube_ausb_003.A12, useNA = "always") # check frequency of response "other"

hchs_ses$prof_edu <- apply(hchs_ses[, c("am_schube_ausb_003.A1", "am_schube_ausb_003.A2", "am_schube_ausb_003.A3", 
                                        "am_schube_ausb_003.A4", "am_schube_ausb_003.A5", "am_schube_ausb_003.A6", 
                                        "am_schube_ausb_003.A7", "am_schube_ausb_003.A8", "am_schube_ausb_003.A9", 
                                        "am_schube_ausb_003.A10", "am_schube_ausb_003.A11")], 
                              1, function(x) {
                                # Convert to numeric, suppressing warnings for NA coercion
                                x_numeric <- suppressWarnings(as.numeric(x))
                                
                                if (all(is.na(x_numeric))) {
                                  return(NA) # Return NA if all variables are NA
                                } else {
                                  return(max(x_numeric, na.rm = TRUE)) # Compute max otherwise
                                }
                              })

table(hchs_ses$prof_edu, useNA = "always") # check result 
```
The frequency table of "am_schube_ausb_003.A12" shows that 498 individuals indicated to have some "other" educational degree. These make up <5% of the sample, so we ignore those and count the next highest educational degree they have checked ("am_schube_ausb_003.A1"-"am_schube_ausb_003.A11"). 

###Profession
variables: am_schube_ausb_023 - am_schube_ausb_030. Take a look at frequency tables original variables. 
```{r}
prof_freq_tables <- list() # create empty list
prof_variables <- c("am_schube_ausb_023", "am_schube_ausb_024", "am_schube_ausb_025", # specify variables with information on profession
                    "am_schube_ausb_026", "am_schube_ausb_027", "am_schube_ausb_028",
                    "am_schube_ausb_029", "am_schube_ausb_030")

for (var_name in prof_variables) { # iterate over variables, create frequency tables 
  prof_freq_table <- table(hchs_ses[[var_name]], useNA = "always")
  prof_freq_tables[[var_name]] <- prof_freq_table
}
print(prof_freq_tables)
```

Unify missing values in all profession variables 
```{r}
recode_prof_variables <- function(column) { # indicate how to recode variables 
  case_when(
    column %in% as.character(1:8) ~ column,
    column %in% c('8888', '9999', '-99', 'NA') ~ NA_character_,
    TRUE ~ NA_character_
  )
}

hchs_ses <- hchs_ses %>% # indicate which variables to recode
  mutate(
    profession = recode_prof_variables(am_schube_ausb_023),
    profession_1 = recode_prof_variables(am_schube_ausb_024),
    profession_2 = recode_prof_variables(am_schube_ausb_025),
    profession_3 = recode_prof_variables(am_schube_ausb_026),
    profession_4 = recode_prof_variables(am_schube_ausb_027),
    profession_5 = recode_prof_variables(am_schube_ausb_028),
    profession_6 = recode_prof_variables(am_schube_ausb_029),
    profession_7 = recode_prof_variables(am_schube_ausb_030))

table(hchs_ses$profession, useNA = "always")
```

Take a look at frequency tables of resulting profession variables 
```{r}
prof_freq_tables_2 <- list() # create empty list
prof_variables_2 <- c("profession", "profession_1", "profession_2", "profession_3", # specify variables with information on profession
                    "profession_4", "profession_5", "profession_6", "profession_7")
for (var_name in prof_variables_2) { # iterate over variables, create frequency tables 
  prof_freq_table_2 <- table(hchs_ses[[var_name]], useNA = "always")
  prof_freq_tables_2[[var_name]] <- prof_freq_table_2
}
print(prof_freq_tables_2)
```

###Income
variables: am_schube_ausb_037 (individual income), am_schube_ausb_038 (household income). Take a look at frequency tables of original income variables.
```{r}
table(hchs_ses$am_schube_ausb_037, useNA = "always")
table(hchs_ses$am_schube_ausb_038, useNA = "always")
```
Variable am_schube_ausb_037 has 3 categories of missing or unclassified values: NA, "8888" (don't want to respond), "9999" (don't know). Classify these values as NAs.
Variable am_schube_ausb_038 has 3 categories of missing or unclassified values: NA, "8888" (don't want to respond), "9999" (don't know). Classify these values as NAs.

```{r}
recode_column <- function(column) {
  case_when(
    column %in% 1:17 ~ as.numeric(column),
    column %in% c(8888, 9999, 66, -99, NA) ~ NA_real_,
    TRUE ~ NA_real_
  )
}

hchs_ses <- hchs_ses %>%
  mutate(
    ind_income = recode_column(am_schube_ausb_037),
    house_income = recode_column(am_schube_ausb_038)
  )
```

Create two variables (individual and household income) with the mean income per category. 
```{r}
class_means <- c(
  `1` = 410,
  `2` = 625,
  `3` = 875,
  `4` = 1125,
  `5` = 1375,
  `6` = 1625,
  `7` = 1875,
  `8` = 2125,
  `9` = 2375,
  `10` = 2750,
  `11` = 3250,
  `12` = 3750,
  `13` = 4250,
  `14` = 4750,
  `15` = 5500,
  `16` = 7000,
  `17` = 9000
)

hchs_ses <- hchs_ses %>%
  mutate(
    ind_income_classmean = case_when(
      ind_income %in% names(class_means) ~ class_means[ind_income],
      is.na(ind_income) ~ NA_real_,
      TRUE ~ NA_real_
    ),
    house_income_classmean = case_when(
      house_income %in% names(class_means) ~ class_means[house_income],
      is.na(house_income) ~ NA_real_,
      TRUE ~ NA_real_
    )
  )

table(hchs_ses$ind_income, useNA = "always")
table(hchs_ses$house_income, useNA = "always")
table(hchs_ses$ind_income_classmean, useNA = "always")
table(hchs_ses$house_income_classmean, useNA = "always")
```

###Household size
variables: sozio_fami_007 (no. of individuals at least 14 years of age in the household including oneself), sozio_fami_006 (no. of individuals in household including oneself)
Take a look at frequency tables of original household size variables
```{r}
table(hchs_ses$sozio_fami_007, useNA = "always")
table(hchs_ses$sozio_fami_006, useNA = "always")
```

Variable sozio_fami_007 and sozio_fami_006 both have 3 categories that need to be reclassified: NA, "8888" (don't want to respond), "9999" (don't know). Classify these values as NAs.
```{r}
hchs_ses <- hchs_ses %>%
  mutate(
    no_adults = case_when(
      sozio_fami_007 %in% as.character(1:18) ~ as.numeric(sozio_fami_007),
      sozio_fami_007 %in% c('8888', '9999', 'NA') ~ NA_real_,
      TRUE ~ as.numeric(sozio_fami_007)
    ),
    no_household = case_when(
      sozio_fami_006 %in% as.character(1:16) ~ as.numeric(sozio_fami_006),
      sozio_fami_006 %in% c('8888', '9999', 'NA') ~ NA_real_,
      TRUE ~ as.numeric(sozio_fami_006)
    )
  )

table(hchs_ses$no_adults, useNA = "always")
table(hchs_ses$no_household, useNA = "always")
```

Number of children in household 
Based on the above information, determine the hypothetical number of children in each household. 
```{r}
hchs_ses <- hchs_ses %>%
  mutate(
    no_children = no_household - no_adults
  )
table(hchs_ses$no_children, useNA = "always") # check if there are any negative values

hchs_ses <- hchs_ses %>% # change negative values to NAs 
  mutate(
    no_children = ifelse(no_children < 0, NA_real_, no_children)
  )
table(hchs_ses$no_children, useNA = "always") 
```

Save preimp_excl_outlier dataframe
```{r, include = FALSE}
write.csv(hchs_ses, file = "hchs_ses_preprocessed_17000.csv", row.names = FALSE)
```

