---
title: "HCHS SES Index determination n=17.000"
output:
  html_document: default
date: "2024-11-16"
authors: A. Klimesch, L. Ascone 
references: ChatGPT was used in the development or adaptation of this code. OpenAI ChatGPT (November 2024 version), https://chat.openai.com/
dataset: initial 17.000 cases of HCHS "HCHS_Data_SPSS_labelled_sozio"
purpose: "this code determines the SES-Index and its subscales for the initial 17.000 HCHS data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load required packages
```{r, results = FALSE, message=FALSE}
library(foreign) 
library(dplyr)
library(haven)
library(psych) # for psych:describe
```

Load dataset 
```{r}
hchs_ses <- read.csv("C:/Users/annek/Documents/PhD/PTSD/HCHS/SES_Index_paper/17000_sample_ses_index/hchs_ses_preprocessed_17000.csv")
```

Education
variables: school_edu, prof_edu
Assign values to the SES-Index subscore education based on the combination of school education and professional education an individual has.
```{r}
hchs_ses <- hchs_ses %>% 
  mutate(
    education = case_when(
      school_edu %in% 1:2 & prof_edu %in% 1:3 ~ 1,
      school_edu %in% 2:3 & prof_edu %in% 4:7 ~ 3,
      school_edu == 3 & prof_edu %in% 1:3 ~ 1.7,
      school_edu %in% 4:6 & prof_edu %in% 1:3 ~ 2.8,
      school_edu %in% 4:6 & prof_edu %in% 4:7 ~ 3.6,
      school_edu %in% 7:9 & prof_edu %in% 1:3 ~ 3.7,
      school_edu %in% 7:9 & prof_edu %in% 4:7 ~ 4.8,
      school_edu %in% 7:9 & prof_edu %in% 8:9 ~ 6.1,
      school_edu %in% 7:9 & prof_edu %in% c(10, 11) ~ 7
    )
  )

table(hchs_ses$education, useNA = "always") # frequencies 
total_n_education <- hchs_ses %>% # check if and how many cases exist with combinations of school_edu and prof_edu that have not been considered by the index classification
  summarise(
    cases_with_values_both = sum(!is.na(school_edu) & !is.na(prof_edu)),
    sum_cases_with_education = sum(!is.na(education))
  )
total_n_education
```

Joblevel and position 
Assign values to the SES-Index subscore job level based on the combination of the occupation and position an individual has.
```{r}
hchs_ses <- hchs_ses %>% 
  mutate(
    job = case_when(
      profession == 1 & profession_1 == 1 ~ 1.1,
      profession == 1 & profession_1 %in% 2:3 ~ 1.0,
      profession == 2 & profession_2 == 1 ~ 6.2,
      profession == 2 & profession_2 == 2 ~ 6.8,
      profession == 2 & profession_2 == 3 ~ 7.0,
      profession == 3 & profession_3 == 1 ~ 3.5,
      profession == 3 & profession_3 == 2 ~ 3.6,
      profession == 3 & profession_3 %in% 3:4 ~ 4.2,
      profession == 4 & profession_4 == 1 ~ 2.9,
      profession == 4 & profession_4 == 2 ~ 4.1,
      profession == 4 & profession_4 == 3 ~ 5.2,
      profession == 4 & profession_4 == 4 ~ 6.4,
      profession == 5 & profession_5 == 1 ~ 2.4, 
      profession == 5 & profession_5 == 2 ~ 3.6,
      profession == 5 & profession_5 == 3 ~ 4.2,
      profession == 5 & profession_5 == 4 ~ 4.7,
      profession == 6 & profession_6 == 1 ~ 1.3,
      profession == 6 & profession_6 %in% 2:3 ~ 2.1,
      profession == 6 & profession_6 == 4 ~ 2.0,
      profession == 6 & profession_6 == 5 ~ 2.4,
      profession == 7 & profession_7 %in% 1:3 ~ 1.0,
      profession == 8 ~ 2.4
    )
  )

table(hchs_ses$education, useNA = "always") # frequencies 
total_n_edu_and_prof <- hchs_ses %>% # check if and how many cases exist with combinations of school_edu and prof_edu and profession that have not been considered by the index classification
  summarise(
    cases_with_values_education_and_profession = sum(!is.na(school_edu) & !is.na(prof_edu) & !is.na(profession)),
    sum_cases_with_profession = sum(!is.na(profession))
  )
total_n_edu_and_prof
```

Income 
variables: no_children, no_adults, house_income_classmean
Determine net-equivalent household income: weight the household income by the number of adults (0.5) and children (0.3) living in the household 
```{r}
hchs_ses <- hchs_ses %>%
  mutate(no_adults_except_subject = no_adults - 1, # substract by 1 because the subjects themselves should not be weighted
         weight_adults = 1 + 0.5 * no_adults_except_subject,
         weight_children = no_children * 0.3,
         weight_netequ_household_inc = weight_adults + weight_children)

table(hchs_ses$weight_netequ_household_inc, useNA = "always") # frequencies

hchs_ses <- hchs_ses %>% 
  mutate(
    netequ_household_inc = case_when(
      weight_netequ_household_inc == 0 ~ house_income_classmean,
      weight_netequ_household_inc > 0 ~  house_income_classmean/weight_netequ_household_inc,
      is.na(weight_netequ_household_inc) ~ NA_real_
      )
    )

table(hchs_ses$netequ_household_inc, useNA = "always")
```

Create 13 categories of netequ_household_inc (based on percentile cut-offs of RKI DEGS study). Assign values to the SES-Index sub score job income based on the net-equivalent-income-quantile the individual belongs to.
```{r}
# Create a vector of percentiles
percentiles <- c(7.6923076923076923076923076923077, 15.384615384615384615384615384615,
                 23.076923076923076923076923076923, 30.769230769230769230769230769231,
                 38.461538461538461538461538461539, 46.153846153846153846153846153846,
                 53.846153846153846153846153846154, 61.538461538461538461538461538462,
                 69.230769230769230769230769230769, 76.923076923076923076923076923077,
                 84.615384615384615384615384615385, 92.307692307692307692307692307692)
quantiles <- percentiles / 100

epsilon <- 1e-5
hchs_ses$netequ_household_inc_jittered <- hchs_ses$netequ_household_inc + runif(length(hchs_ses$netequ_household_inc), -epsilon, epsilon) # jitter values so quantiles can be created (there are many similar values, therefore, quantiles would not be the same size without jittering)

quantile_results <- quantile(hchs_ses$netequ_household_inc_jittered, probs = quantiles, na.rm = TRUE) # create quantiles 

breaks <- c(-Inf, quantile_results, Inf) # create breaks for cut function

hchs_ses <- hchs_ses %>% # assign values to values depending on quantile 
  mutate(
    income = cut(netequ_household_inc_jittered, breaks = breaks, labels = c(1.0, 1.5, 2.0, 2.5, 3.0, 3.5, 4.0, 4.5, 5.0, 5.5, 6.0, 6.5, 7.0), include.lowest = TRUE)
  )

table(hchs_ses$income)

total_n_edu_prof_inc <- hchs_ses %>% # check if and how many cases exist with combinations of school_edu and prof_edu and profession that have not been considered by the index classification
  summarise(
    cases_with_values_edu_and_prof_and_inc = sum(!is.na(education) & !is.na(profession) & !is.na(income)),
    sum_cases_with_income = sum(!is.na(income))
  )
total_n_edu_prof_inc
```

SES-Index
Determine the index based on the index sub scores 
```{r}
hchs_ses <- hchs_ses %>%
  mutate(
    education = as.numeric(as.character(education)),
    job = as.numeric(as.character(job)),
    income = as.numeric(as.character(income)),
    ses_index = education + job + income
  )

table(hchs_ses$ses_index)
hist(hchs_ses$ses_index)
sum(!is.na(hchs_ses$ses_index)) # determine how many ses_index values were determined
```

Save ses-index data frame
```{r, include = FALSE}
write.csv(hchs_ses, file = "hchs_ses_index_preimp_17000.csv", row.names = FALSE)
```
