---
title: "HCHS SES Index Imputation n=17.000"
output: html_document
date: "2024-11-18"
author: "Anne Klimesch"
references: ChatGPT was used in the development or adaptation of this code. OpenAI ChatGPT (November 2024 version), https://chat.openai.com/; The following website was used for the imputation procedure https://rpubs.com/kaz_yos/mice-exclude
note: |
  This script prepares the dataset n=17.000 for imputation of the HCHS SES-Index and performs a multiple imputation. The output of the script are two data files "hchs_index_postimp_17000.csv" and "hchs_index_postimp_excl_outlier_17000.csv".
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load required packages
```{r, include=FALSE}
library(dplyr)
library(stringr)  #extract parts of a variable for each case of this variable; organize columns when reading the mri txt files; to remove spaces from values
library(haven)    #to import spss data files
library(foreign)
library(naniar)   #for missing variable analysis 
library(nortest)  #for missing variable analysis 
library(stats)    #for missing variable analysis 
library(kableExtra)#for missing data analysis
library(ggplot2)  #for histograms
library(gridExtra)#for presentation of histograms 
library(finalfit) #for visualization of missing data patterns 
library(psych)    #for descriptives and intercorrelation matrix
library(sjPlot)   #for plotting descriptives
library(stargazer)#for plotting descriptives
library(ggplot2)  #for plotting  
library(gridExtra)#for plotting 
library(visdat) #for visualization of missing data pattern for many variables at the same time
library(apaTables)
library("Hmisc")
library(tidyverse)
library(mice)     #for imputation
library(ggplot2) # for comparison of pre- and post-imp distribution 
library(finalfit) # for missing plot function
```
```

###IMPORTING HCHS-SES-INDEX DATA
#Import previously determined hchs ses-index 
```{r}
ses_data <- read.csv("C:/Users/annek/Documents/PhD/PTSD/HCHS/SES_Index_paper/17000_sample_ses_index/hchs_ses_index_preimp_17000.csv")
```

###PREPARATION SES DATA
```{r}
ses_data_subset <- ses_data[, c("DisclosureID", #choose subset of variables
                                "HCH_SVAGE0001", 
                                "HCH_SVSEX0001", 
                                "job", 
                                "income", 
                                "education",
                                "ses_index"
                                )]

colnames_ses_subset <- colnames(ses_data_subset) #save existing names to character vector
ses_cols_to_replace <- c("DisclosureID", "HCH_SVAGE0001", "HCH_SVSEX0001")
new_names <- c("pseudonym", "age", "sex")
replace_indices <- which(colnames_ses_subset %in% ses_cols_to_replace)
colnames(ses_data_subset)[replace_indices] <- new_names
print(colnames(ses_data_subset))
```

###IDENTIFY MISSING DATA (PATTERN)
Create missing plots to get an overview of the missing data pattern
```{r}
ses_missing_analysis <- ses_data_subset[, c("age", "sex", "job", "income", "education", "ses_index")]
ses_missing_analysis %>% #save variables of interest as new data frame and create a missing plot with them
  missing_plot(title = "Missing values plot of initial SES and questionnaire data (n = 17.000)")

explanatory = c("job", "income", "education", "ses_index")
dependent = "vf_phq9_001"
ses_data_subset %>% # choose subset of variables and look at their missing data pattern
  missing_pattern(dependent, explanatory)

table_output <- miss_var_summary(ses_missing_analysis) # check for % of missing values per variable
table_html <- kable(table_output, format = "html") %>% # convert the table to a html table
  kable_styling()
writeLines(table_html, "missing_summary.html") #save html
```

###PRE-IMPUTATION DATASET
#Create variables which indicate missing values for each case in each ses variable. The variables will be used in multiple imputation, however, these variables have been imputed in another study so we do not want to impute the missing values again but keep the NAs. Therefore, I save these NAs now, to later recreate them.
```{r}
check_na <- c("job", "education", "income", "ses_index") # Define the variables to check for NA values
for (variable in check_na) { # iterate over the variables 
  na_variable <- paste0(variable, "_na") # create a new variable with _na suffix for each variable
  ses_data_subset[[na_variable]] <- ifelse(is.na(ses_data_subset[[variable]]), 0, 1) # check for NAs in the variables and assign 0=NA or 1=no_NA accordingly
}
table(ses_data_subset$education_na) # NAs n=2340
table(ses_data_subset$job_na) # NAs n=3428
table(ses_data_subset$income_na) # NAs = 6586
```

MULTIPLE IMPUTATION (https://rpubs.com/kaz_yos/mice-exclude)
```{r, results='hide'}
allVars <- names(ses_data_subset) # save all variable names in the data frame in a list

missVars <- names(ses_data_subset)[colSums(is.na(ses_data_subset)) > 0] # save names of variables with NAs in a list

predictorMatrix <- matrix(0, ncol = length(allVars), nrow = length(allVars)) # create mice predictor matrix (Rows correspond to target variables;  A value of '1' means that the column variable is used as a predictor for the target variable (in the rows)) 
rownames(predictorMatrix) <- allVars
colnames(predictorMatrix) <- allVars

imputerVars <- c("age", "sex", "job", "income", "education") # variables that will inform the imputation

imputerVars <- intersect(unique(imputerVars), allVars) #  remove duplicate names in "imputer_vars"; take the two lists and keep only those variables that exist in both
imputerVars
imputerMatrix <- predictorMatrix # create a copy of predictor_matrix
imputerMatrix[,imputerVars] <- 1 #  set the imputer_vars in the predictor_matrix to 1 to indicate that they are predictors
imputerMatrix

imputedVars <- intersect(unique(imputerVars), missVars) # remove duplicate names in "imputer_vars"; take the two lists and keep only those variables that exist in both
imputedVars # check which variables are left; age and sex were removed because they do not contain NAs 
imputedMatrix <- predictorMatrix
imputedMatrix[imputedVars,] <- 1 # set the imputed_vars in the imputed_matrix to 1 to indicate that they are targets
imputedMatrix

predictorMatrix <- imputerMatrix * imputedMatrix # construct a full predictor matrix (rows: imputed variables; cols: imputer variables)
diag(predictorMatrix) <- 0 # diagonals must be zeros because a variable cannot impute itself
predictorMatrix

set.seed(156)
dryMice <- mice(data = ses_data_subset, m = 1, predictorMatrix = predictorMatrix, maxit = 0, seed = 156) # perform a pilot-imputation without imputing NAs (maxit = 0)
dryMice$method # get the imputation method assigned to each variable (pmm = predictive mean matching, logreg = logistic regression, "" = no imputation planned~~)
predictorMatrix <- dryMice$predictorMatrix # update predictor matrix
options(max.print = 9999, width = 200) #adjust max.print value to make whole imputation matrix visible 
print(predictorMatrix) # print whole imputation matrix
imputerVars <- colnames(predictorMatrix)[colSums(predictorMatrix) > 0] #Imputers (non-zero columns of predictorMatrix)
imputerVars 
imputedVars <- rownames(predictorMatrix)[rowSums(predictorMatrix) > 0] #Imputed (non-zero rows of predictorMatrix)
imputedVars 
setdiff(imputerVars, imputedVars) # imputers that are complete
intersect(imputerVars, imputedVars) # imputers with NAs
setdiff(imputedVars, imputerVars) # imputed-only variables without being imputers
setdiff(missVars, imputedVars) # variables with NAs that are not imputed
predictorMatrix[rowSums(predictorMatrix) > 0, colSums(predictorMatrix) > 0] # Relevant part of predictorMatrix

set.seed(4260)
miceout <- mice(data = ses_data_subset, m = 5, predictorMatrix = predictorMatrix, method = dryMice$method, maxit = 50, seed = 4260)

completedDatasets <- vector("list", length = 5) # Create an empty list to store the completed datasets
for (i in 1:5) { # store each completed dataset in the list
  completedDatasets[[i]] <- complete(miceout, action = i)
}

actuallyImputedVars <- vector("list", length = 5) # get information on variables actually imputed for each dataset
for (i in 1:5) {
  actuallyImputedVars[[i]] <- setdiff(names(ses_data_subset)[colSums(is.na(ses_data_subset)) > 0], 
                                      names(completedDatasets[[i]])[colSums(is.na(completedDatasets[[i]])) > 0]) 
} # Variables that no longer have NAs after imputation

unexpectedlyImputedVars <- vector("list", length = 5) # Get variables that were unexpectedly imputed for each dataset
for (i in 1:5) {
  unexpectedlyImputedVars[[i]] <- setdiff(actuallyImputedVars[[i]], imputedVars)
}

plannedNotImputedVars <- vector("list", length = 5) # Get variables that were planned to be but not imputed for each dataset
for (i in 1:5) {
  plannedNotImputedVars[[i]] <- setdiff(imputedVars, actuallyImputedVars[[i]])
}

stillMissingVars <- vector("list", length = 5) # Get still missing variables for each dataset
for (i in 1:5) {
  stillMissingVars[[i]] <- names(completedDatasets[[i]])[colSums(is.na(completedDatasets[[i]])) > 0]
}

for (i in 1:5) { # Print results for each dataset separately
  print(actuallyImputedVars[[i]]) #Variables actually imputed
  print(unexpectedlyImputedVars[[i]]) #Variables unexpectedly imputed
  print(plannedNotImputedVars[[i]]) #Variables planned but not imputed
  print(stillMissingVars[[i]]) #Still missing variables
}

postimp_1 <- complete(miceout,1)
postimp_2 <- complete(miceout,2)
postimp_3 <- complete(miceout,3)
postimp_4 <- complete(miceout,4)
postimp_5 <- complete(miceout,5)

postimp <-   complete(miceout) # merge imputation sets 
str(postimp)

imputation_methods <- miceout$method # check if the correct methods for imputation have been used 
print(imputation_methods)
nrow(postimp) # n=16411
```

Create SES-Index
```{r}
postimp <- postimp %>%
  mutate(
    education = as.numeric(as.character(education)),
    job = as.numeric(as.character(job)),
    income = as.numeric(as.character(income)),
    ses_index = education + job + income
  )

table(postimp$ses_index)
hist(postimp$ses_index)
sum(!is.na(postimp$ses_index)) # determine how many ses_index values were determined
```

##Recreate NAs
#1) Set those cases to NA, which had an NA in all subscores of the SES-Index. This way we keep only those imputations that are based on at least one subscore. 
#2) Check the number of NAs on the ses_index variable
#3) Create a postimp_no_na variable, excluding all cases that have an NA on the ses_index 
```{r}
postimp <- postimp %>%
  mutate(
    ses_index = if_else(education_na == 0 & job_na == 0 & income_na == 0, NA_real_, ses_index)
  )

# Then, update education, job, and income based on the updated ses_index
postimp <- postimp %>%
  mutate(
    education = if_else(is.na(ses_index) & education_na == 0, NA_real_, education),
    job = if_else(is.na(ses_index) & job_na == 0, NA_real_, job),
    income = if_else(is.na(ses_index) & income_na == 0, NA_real_, income)
  )
  
sum(!is.na(postimp$ses_index))

postimp_no_na <- postimp[!is.na(postimp$ses_index), ]
```

COMPARE DISTRIBUTION BEFORE AND AFTER IMPUTATION (chat gpt)  
```{r}
plot_folder <- "imputation_quality_check_density_plots" # check if folder exists and if not create it
if(!dir.exists(plot_folder)) {
  dir.create(plot_folder)
}

plots <- list() # Initialize an empty list to store plots
for(var in imputedVars) { # Loop through each variable and create a density plot
  plots[[var]] <- ggplot() +
    geom_density(data = ses_data_subset, aes_string(x = var, y = "..density.."), fill = "blue", alpha = 0.4) +
    geom_density(data = postimp, aes_string(x = var, y = "..density.."), fill = "red", alpha = 0.4) +
    labs(title = paste("Density Plot of", var, "Pre and Post Imputation"),
         x = var,
         y = "Density") +
    theme_minimal()
  file_path <- paste0(plot_folder, "/", "density_plot_", var, ".png") # Define the file path within the new folder
  ggsave(file_path, plot = plots[[var]], width = 10, height = 8) # Save the plot to the specified file path
}

vars_ses <- c("income", "education", "job", "ses_index") # choose variables 
for(var in vars_ses) { # create tables of the chosen variables and compare pre-post imputation for each of them 
  cat("Frequency table for:", var, " - Pre-Imputation\n")
  print(summary(ses_data_subset[[var]], useNA = "ifany"))
  cat("\n") # adds an empty line for readability
  
  cat("Frequency table for:", var, " - Post-Imputation\n")
  print(summary(postimp[[var]], useNA = "ifany"))
  cat("\n\n") # adds two empty lines for readability between variable comparisons
}
hist(postimp$ses_index, main = "Histogram of ses index after imputation")
hist(ses_data_subset$ses_index, main = "Histogram of ses index before imputation")
```

For clarification, change variable names
```{r}
postimp <- postimp %>%
  rename(
    education_computed_based_on_raw_data = education_na,
    job_computed_based_on_raw_data = job_na,
    income_computed_based_on_raw_data = income_na,
    index_computed_based_on_raw_data = ses_index_na
  )
```

#Save as csv files
I call the data frame "17000" even though it acutally only includes 16411. This is, because it was supposed to be the "n=17.000" subsample of the HCHS study. 
```{r}
write.csv(postimp, file = "hchs_index_postimp_17000.csv", row.names = FALSE) # Save postimp dataframe

#Save pseudonym values of those cases that have ses_index values (n=15292)
length(postimp_no_na$pseudonym) # check for number of cases 
sum(is.na(postimp_no_na$pseudonym)) # check for NAs
pseudonym_values_15292 <- postimp_no_na$pseudonym
values_string_15292 <- paste(pseudonym_values_15292, collapse = "\n") # Create a string with values separated by newline character
write(values_string_15292, "pseudonym_values_postimp_15292.txt")  # Write the string to a text file
```
